%************************************************
\chapter{Entrance and Vesting}\label{ch:entrance-vesting}
%************************************************
\instr{The priest enters the altar (the deacon remains at the ambo), performs two full bows or bows from the waist before the Holy Table, kisses it, performs a third bow, puts on his epitrachilion, takes the covering off of the Holy Table, exits the altar via the north door and goes down before the ambo.}

\instr{The priest and deacon make the sign of the cross and bow from the waist three times.}

\dc{Bless, Master}

\pr{Blessed is our God, always, now and ever, and unto the ages of ages.}

\dc{Amen. O Heavenly King\ldots Holy God\ldots Most Holy Trinity\ldots Our Father\ldots}

\pr{For Thine is the kingdom, and the power, and the glory of the Father, and of the Son, and of the Holy Spirit, now and ever, and unto the ages of ages.}

\dc{Amen.}

\prdc{Have mercy on us, O Lord\ldots Glory\ldots O Lord, have mercy on us\ldots Both now\ldots The door of compassion\ldots}

\instr{The priest and deacon approach the icon of Christ the Savior and make the sign of the cross with bows.}

\prdc{We venerate Thine immaculate icon, O Good One\ldots}

\instr{Making the sign of the cross and bowing before the icon of the Savior, they kiss it.}

\instr{They then approach the icon of the most holy Theotokos and make the sign of the cross with bows.}

\prdc{As thou art a wellspring of compassion\ldots}

\instr{Making the sign of the cross and bowing before the icon of the Theotokos, they kiss it.}

\instr{Then they stand again before the Royal Gates and bow their heads.}

\pr{O Lord, send down\ldots}

\instr{The priest and deacon then bow to each other, turn around and bow to the faithful, and then bow to anyone on the left and right kliros, saying}

\prdc{Forgive and bless, fathers and brethren.}

\instr{They enter the altar via the south door, reading part of the 5th Psalm:}

\prdc{I shall go into Thy house\ldots}

\instr{The deacon, crossing himself, makes two bows (either full or from the waist) before the Holy Table, kisses there the Gospel, Cross, and the Holy Table itself, bows a third time, and then makes one more bow towards the High Place. He then goes to get vested.}

%************************************************
\section{Vesting}\label{sec:vesting}
%************************************************
\instr{The deacon folds his sticharion and places on top of it the orarion and cuffs, then crosses himself and makes three bows toward the High Place, saying aloud to himself each time:}

\dc{God, cleanse me, a sinner, and have mercy on me. \numtimes{3}}

\instr{He then approaches the priest, bows his head and says:}

\dc{Bless, master, the sticharion with the orarion.}

\pr{Blessed is our God\ldots}

\dc{Amen.}

\instr{The deacon kisses the priests hand and the cross on the back of his sticharion, and then reverently puts on his \textbf{sticharion}, saying:}

\dc{My soul shall rejoice in the Lord\ldots}

\instr{He then takes the \textbf{orarion} and, kissing the cross on it, puts it on his left shoulder.}

\instr{He then takes one \textbf{cuff}, kisses the cross on it, and puts it on his \textbf{right} arm, saying:}

\dc{Thy right hand, O Lord\ldots}

\instr{He then takes the other \textbf{cuff}, kisses the cross on it, and puts it on his \textbf{left} arm while saying:}

\dc{Thy hands have made me\ldots}

\instr{After vesting, the deacon (and separately the priest) washes his hands, saying the 25th Psalm:}

\dc{I will wash my hands in innocence\ldots...}

\instr{At this point, the deacon should prepare everything needed for Holy Communion: remove any covers from the Holy Table and the Table of Oblation, light candles and lamps on them, place the diskos (on the left) and the chalice (on the right) of the Table of Oblation as well as the spear, spoon, veils/covers, the aer, prosphora(s), wine, bowl, pitcher, plate, and sponge.}
