%************************************************
\chapter{Proskomide}\label{ch:proskomide}
% ************************************************
\instr{The priest and deacon approach the Table of Oblation and bow three times with the words:}

\prdc{God, cleanse me, a sinner, and have mercy on me. \numtimes{2}\\
  Thou has delivered us from the curse by Thy precious Blood\ldots}

\instr{The deacon points to the holy implements and says:}

\dc{Bless, master.}

\pr{Blessed is our God\ldots}

\dc{Amen.}

\instr{After the priest says ``Blessed is our God\ldots,'' he takes the prosphora (Greek style, the only prosphora; Russian style, the prosphora with the sign of the cross that will be used as the Lamb) in his left hand and uses his right hand to make the sign of the cross with the spear over the seal of the prosphora three times, each time saying:}

\pr{In remembrance of our Lord, and God, and Savior Jesus Christ.}

\instr{Then the deacon begins with orarion raised:}

\dc{Let us pray to the Lord.}

\instr{The priest cuts the Lamb on its right side (priest's left) with the words:}

\pr{Like a lamb\ldots}

\dc{Let us pray to the Lord.}

\instr{The priest cuts the Lamb on its left side (priest's right) with the words:}

\pr{And as Lamb blameless\ldots}

\dc{Let us pray to the Lord.}

\instr{The priest cuts the top of the Lamb with the words:}

\pr{In His humility\ldots}

\dc{Let us pray to the Lord.}

\instr{The priest cuts the bottom of the Lamb with the words:}

\pr{Who shall declare\ldots}

\dc{Take out, master.}

\instr{The priest thrusts the spear into the bottom of the Lamb (Greek style, coming in on the right side (priest's left) of the Lamb) and pries out the Lamb, saying the words:}

\pr{For His life\ldots}

\instr{The priest places the Lamb with the seal face-down on the diskos.}

\dc{Sacrifice, master.}

\instr{The priest makes a deep cross-wise incision on the bottom (non-seal) side of the Lamb down to (but not through) the seal, saying:}

\pr{Sacrificed is the Lamb of God\ldots}

\instr{The priest flips the Lamb so the seal is now on top.}

\dc{Pierce, master.}

\instr{The priest pierces the Lamb with the lance on its right side (priest's left) saying:}

\pr{One of the soldiers\ldots}

\instr{The deacon takes the two small pitchers, one of water and one of wine, saying:}

\dc{Bless, master, the holy union.}

\instr{The priest blesses.}

\instr{The deacon pours water and wine simultaneously into the chalice. At this point, the deacon may leave to attend to other things until it is time to cover the gifts.}

\instr{If using multiple prosphoras, the priest takes the second one, if possible one made with a Theotokos seal. If using a single Greek-style prosphora, a large triangle is already printed to the right of where the Lamb was.}

Further priest instructions to be included at a later date.

%************************************************
\section{Covering of the Gifts}\label{sec:covering-gifts}
%************************************************

\instr{After the priest has finished all of the comemmorations, it is time to finish the service of Proskomide and to cover the gifts.}

\instr{The deacon takes the censer, puts incense in it, carries it to the priest and raising it up says:}

\dc{Bless, master, the censer. Let us pray to the Lord.}

\pr{Incense we offer\ldots}

\instr{The deacon opens the censer and, using his orarion to shield his hand from the heat, holds it open at the level of the Table of Oblation for the priest to access, saying:}

\dc{Let us pray to the Lord.}

\instr{The priest takes the star and makes the sign of the cross with it over the smoke from the censer, then places it open over the Lamb on the diskos while saying:}

\pr{And the star came\ldots}

\dc{Let us pray to the Lord.}

\instr{The priest takes the first small veil, makes the sign of the cross with it over the smoking censer, places it on the diskos and says:}

\pr{The Lord is King\ldots}

\dc{Let us pray to the Lord. Cover, master.}

\instr{The priest takes the second small veil, makes the sign of hte cross with it over the smoking censer, places it on the chalice and says:}

\pr{Cover the heavens\ldots}

\instr{The deacon holds the censer normally, keeping the chamber still slightly open and says:}

\dc{Let us pray to the Lord. Cover, master.}

\instr{The priest wraps the aer around the censer, then takes it and covers both the already-covered diskos and chalice, saying:}

\pr{Cover us with the shelter\ldots}

\instr{The priest takes the censer from the deacon, censes the Table of Oblation while having the following exchange three times:}

\pr{Blessed is our God, Who is thus well pleased, glory to Thee.}

\dc{Always, now and ever, and unto the ages of ages. Amen.}

\dc{For the Precious Gifts set for, let us pray to the Lord.}

\pr{O God, our God\ldots Glory to Thee, O Christ God, our Hope, glory to Thee.}

\dc{Glory to the Father, and to the Son, and to the Holy Spirit. Both now, and ever, and unto the ages of ages. Amen. Lord, have mercy. \numtimes{3} Father, bless.}

\pr{May Christ our true God\ldots}

\dc{Amen.}

\instr{The priest then censes the deacon, hands off the censer to the deacon who in return censes the priest.}

%************************************************
\section{Censing at the Hours}\label{sec:censing-at-hours}
% ************************************************
\instr{\marginpar{For the remainder of this document, text in this color will signify instructions meant only for two or more deacons coserving.} The deacon \subinstr{if two deacons serve, then both; if three or more, then the 2nd and the 3rd} goes to the front of the Holy Table and opens the curtain behind the Royal Doors.}

\instr{The deacon censes the Holy Table on all four sides beginning with the west (front) side of the Holy Table and proceeding counter-clockwise \subinstr{three sides, each his own half symmetrically}, saying to himself the following troparion in parts at each side of the Holy Table as follows:}

\dc{\emph{(west)} In the grave bodily, but in Hades with Thy soul as God, \emph{(south)} in Paradise with the thief, \emph{(east)} and on the throne with the Father and the Spirit, \emph{(south)} wast Thou Who fillest all things, O Christ the Inexpressible.}

\instr{The deacon then begins the 50th Psalm, which he continues until the end of the censing. After finishing the Holy Table, the deacon goes to cense the Table of Oblation (via the High Place if on the right side of the altar), then the High Place and the right (south) side of the altar, followed by the left (north) side of the altar \subinstr{each his own side, simultaneously}. Standing near the north door, the deacon then censes the cross on/behind the Holy Table, then goes to the High Place and from their censes the priests and servers.}

\instr{The deacon then exits the altar via the north door \subinstr{each via his own door} and goes to the ambo. He censes the icon of the Mystical Supper on the Holy Doors, the icon of the Savior and the rest of the icons on the right (south) side of the iconostasis, followed by the icon of the Theotokos and the rest of the icons on the left (north) side of the iconostasis \subinstr{each deacon censes his side of the iconostasis, simultaneously}. Returning to the ambo, the deacon then censes any priests on either side of the kliros, the choir, and the people \subinstr{each his own side of the church, simultaneously}.}

\instr{Then the deacon goes down to the festal icon in the center of the nave \subinstr{each via his own side}, censes it and then all the icons in the church clock-wise \subinstr{each his own side of the church, simultaneously}. He then goes back up to the ambo, censes the icon of the Mystical Supper on the Royal Gates, the icon of the Savior, the icon of the Theotokos \subinstr{each his own side}, enters the altar via the south door \subinstr{via both doors}, censes the front of the Holy Table, goes to the High Place \subinstr{each via his own side}, censes the priest (and if the priest is absent from the altar, he censes the High Place instead) \subinstr{then each other}, makes the sign of the cros and bows towards the high Place, the priest, \subinstr{then each other} and then hands off the censer.}
