\documentclass[
  14pt,
  %% draft,
  letterpaper,
  openright,
  twocolumn,
  twoside]{memoir}

\input{preamble}

\begin{document}

\chapter*{Matins}
%% \chapterprecis{As served on most Sundays}

\priest\ Blessed is our God, always, now and ever, and unto the ages of ages.

\reader\ Amen.

Glory to Thee, our God, glory to Thee.

O Heavenly King, Comforter, Spirit of Truth, Who art everywhere present and fillest all things, Treasury of good things and Giver of life: Come and dwell in us, and cleanse us of all impurity, and save our souls, O Good One.

Holy God, Holy Mighty, Holy Immortal, have mercy on us. \thrice\

\glory\ \bothnow\

O Most Holy Trinity, have mercy on us. O Lord, blot out our sins. O Master, pardon our iniquities. O Holy One, visit and heal our infirmities for Thy name's sake.

\lhm\ \thrice

\glory\ \bothnow

Our Father, Who art in the Heavens, hallowed be Thy Name. Thy Kingdom come, Thy will be done, on earth as it is in Heaven. Give us this day our daily bread, and forgive us our debts, as we forgive our debtors; and lead us not into temptation, but deliver us from the evil one.

\priest\ For Thine is the kingdom and the power, and the glory: of the Father and of the Son, and of the Holy Spirit, now and ever, and unto the ages of ages.

\reader\ Amen.

Save, O Lord, Thy people, and bless Thine inheritance; grant Thou unto Orthodox Christians victory over enemies; and by the power of Thy Cross do Thou preserve Thy commonwealth.

\glory\

O Thou Who wast lifted up willingly on the Cross, bestow Thy mercies upon the new community named after Thee, O Christ God; gladden with Thy power the Orthodox Christians, granting them victory over enemies; may they have as Thy help the weapon of peace, the invincible trophy.

\bothnow\

O awesome intercession that cannot be put to shame, O good one, disdain not our prayer; O all-hymned Theotokos, establish the commonwealth of the Orthodox, save the Orthodox Christians, and grant unto them victory from heaven, for thou didst bring forth God, O thou only blessed one.

\note{Ektenia. ``\lhm'' three times at each response.}

\reader\ Amen. In the name of the Lord, Father, bless.

\priest\ Glory to the holy, and consubstantial, and life-creating, and indivisible Trinity, always, now and ever, and unto the ages of ages.

\reader\ Amen.

\section*{The Six Psalms}

\note{And we begin the Six Psalms, with each one listening with silence and compunction. And the brother (sister) who readeth with reverence and fear of God, without crosses or bows during the entirety of the reading and with simple voice saith:}

Glory to God in the highest, and on earth peace, good will among men. \thrice\

O Lord, Thou shalt open my lips, and my mouth shall declare Thy praise. \twice\

\centernote{Psalm 3}

O Lord, why are they multiplied that afflict me? Many rise up against me. Many say unto my soul: There is no salvation for him in his God. But Thou, O Lord, art my helper, my glory, and the lifter up of my head. I cried unto the Lord with my voice, and He heard me out of His holy mountain. I laid me down and slept; I awoke, for the Lord will help me. I will not be afraid of ten thousands of people that set themselves against me round about. Arise, O Lord, save me, O my God, for Thou hast smitten all who without cause are mine enemies; the teeth of sinners hast Thou broken. Salvation is of the Lord, and Thy blessing is upon Thy people.

I laid me down and slept; I awoke, for the Lord will help me.

\centernote{Psalm 37}

O Lord, rebuke me not in Thine anger, nor chasten me in Thy wrath. For Thine arrows are fastened in me, and Thou hast laid Thy hand heavily upon me. There is no healing in my flesh in the face of Thy wrath; and there is no peace in my bones in the face of my sins. For mine iniquities are risen higher than my head; as a heavy burden have they pressed heavily upon me. My bruises are become noisome and corrupt in the face of my folly. I have been wretched and utterly bowed down until the end; all the day long I went with downcast face. For my loins are filled with mockings, and there is no healing in my flesh. I am afflicted and humbled exceedingly, I have roared from the groaning of my heart. O Lord, before Thee is all my desire, and my groaning is not hid from Thee. My heart is troubled, my strength hath failed me; and the light of mine eyes, even this is not with me. My friends and my neighbors drew nigh over against me and stood, and my nearest of kin stood afar off. And they that sought after my soul used violence; and they that sought evils for me spake vain things, and craftinesses all the day long did they meditate. But as for me, like a deaf man I heard them not, and was as a speechless man that openeth not his mouth. And I became as a man that heareth not, and that hath in his mouth no reproofs. For in Thee have I hoped, O Lord; Thou wilt hearken unto me, O Lord my God. For I said: Let never mine enemies rejoice over me; yea, when my feet were shaken, those men spake boastful words against me. For I am ready for scourges, and my sorrow is continually before me. For I will declare mine iniquity, and I will take heed concerning my sin. But mine enemies live and are made stronger than I, and they that hated me unjustly are multiplied. They that render me evil for good slandered me, because I pursued goodness. Forsake me not, O Lord my God, depart not from me. Be attentive unto my help, O Lord of my salvation.

Forsake me not, O Lord my God, depart not from me. Be attentive unto my help, O Lord of my salvation.

\centernote{Psalm 62}

O God, my God, unto Thee I rise early at dawn. My soul hath thirsted for Thee; how often hath my flesh longed after Thee in a land barren and untrodden and unwatered. So in the sanctuary have I appeared before Thee to see Thy power and Thy glory. For Thy mercy is better than lives; my lips shall praise Thee. So shall I bless Thee in my life, and in Thy name will I lift up my hands. As with marrow and fatness let my soul be filled, and with lips of rejoicing shall my mouth praise Thee. If I remembered Thee on my bed, at the dawn I meditated on Thee. For Thou art become my helper; in the shelter of Thy wings will I rejoice. My soul hath cleaved after Thee, Thy right hand hath been quick to help me. But as for these, in vain have they sought after my soul; they shall go into the nethermost parts of the earth, they shall be surrendered unto the edge of the sword; portions for foxes shall they be. But the king shall be glad in God, everyone shall be praised that sweareth by Him; for the mouth of them is stopped that speak unjust things.

At the dawn I meditated on Thee. For Thou art become my helper; in the shelter of Thy wings will I rejoice.  My soul hath cleaved after Thee, Thy right hand hath been quick to help me.

\glory\ \bothnow\

\alleluia\ \thrice\

\lhm\ \thrice\

\glory\ \bothnow\

\centernote{Psalm 87}

O Lord God of my salvation, by day have I cried and by night before Thee. Let my prayer come before Thee, bow down Thine ear unto my supplication. For filled with evils is my soul, and my life unto hades hath drawn nigh. I am counted with them that go down into the pit; I am become as a man without help, free among the dead. Like the bodies of the slain that sleep in the grave, whom Thou rememberest no more, and they are cut off from Thy hand. They laid me in the lowest pit, in darkness and in the shadow of death. Against me is Thine anger made strong, and all Thy billows hast Thou brought upon me. Thou hast removed my friends afar from me; they have made me an abomination unto themselves. I have been delivered up, and have not come forth; mine eyes are grown weak from poverty. I have cried unto Thee, O Lord, the whole day long; I have stretched out my hands unto Thee. Nay, for the dead wilt Thou work wonders? Or shall physicians raise them up that they may give thanks unto Thee? Nay, shall any in the grave tell of Thy mercy, and of Thy truth in that destruction? Nay, shall Thy wonders be known in that darkness, and Thy righteousness in that land that is forgotten? But as for me, unto Thee, O Lord, have I cried; and in the morning shall my prayer come before Thee. Wherefore, O Lord, dost Thou cast off my soul and turnest Thy face away from me? A poor man am I, and in troubles from my youth; yea, having been exalted, I was humbled and brought to distress. Thy furies have passed upon me, and Thy terrors have sorely troubled me. They came round about me like water, all the day long they compassed me about together. Thou hast removed afar from me friend and neighbor, and mine acquaintances because of my misery.

O Lord God of my salvation, by day have I cried and by night before Thee. Let my prayer come before Thee, bow down Thine ear unto my supplication.

\centernote{Psalm 102}

Bless the Lord, O my soul, and all that is within me bless His holy name. Bless the Lord, O my soul, and forget not all that He hath done for thee, Who is gracious unto all thine iniquities, Who healeth all thine infirmities, Who redeemeth thy life from corruption, Who crowneth thee with mercy and compassion, Who fulfilleth thy desire with good things; thy youth shall be renewed as the eagle's. The Lord performeth deeds of mercy, and executeth judgment for all them that are wronged. He hath made His ways known unto Moses, unto the sons of Israel the things that He hath willed. Compassionate and merciful is the Lord, longsuffering and plenteous in mercy; not unto the end will He be angered, neither unto eternity will He be wroth. Not according to our iniquities hath He dealt with us, neither according to our sins hath He rewarded us. For according to the height of heaven from the earth, the Lord hath made His mercy to prevail over them that fear Him. As far as the east is from the west, so far hath He removed our iniquities from us. Like as a father hath compassion upon his sons, so hath the Lord had compassion upon them that fear Him; for He knoweth whereof we are made, He hath remembered that we are dust. As for man, his days are as the grass; as a flower of the field, so shall he blossom forth. For when the wind is passed over it, then it shall be gone, and no longer will it know the place thereof. But the mercy of the Lord is from eternity, even unto eternity, upon them that fear Him. And His righteousness is upon sons of sons, upon them that keep His testament and remember His commandments to do them. The Lord in heaven hath prepared His throne, and His kingdom ruleth over all. Bless the Lord, all ye His angels, mighty in strength, that perform His word, to hear the voice of His words. Bless the Lord, all ye His hosts, His ministers that do His will. Bless the Lord, all ye His works, in every place of His dominion. Bless the Lord, O my soul.

In every place of His dominion, bless the Lord, O my soul.

\centernote{Psalm 142}

O Lord, hear my prayer, give ear unto my supplication in Thy truth; hearken unto me in Thy righteousness. And enter not into judgment with Thy servant, for in Thy sight shall no man living be justified. For the enemy hath persecuted my soul; he hath humbled my life down to the earth. He hath sat me in darkness as those that have been long dead, and my spirit within me is become despondent; within me my heart is troubled. I remembered days of old, I meditated on all Thy works, I pondered on the creations of Thy hands. I stretched forth my hands unto Thee; my soul thirsteth after thee like a waterless land. Quickly hear me, O Lord; my spirit hath fainted away. Turn not Thy face away from me, lest I be like unto them that go down into the pit. Cause me to hear Thy mercy in the morning; for in Thee have I put my hope. cause me to know, O Lord, the way wherein I should walk; for unto Thee have I lifted up my soul. Rescue me from mine enemies, O Lord; unto Thee have I fled for refuge. Teach me to do Thy will, for Thou art my God. Thy good Spirit shall lead me in the land of uprightness; for Thy name's sake, O Lord, shalt Thou quicken me. In Thy righteousness shalt Thou bring my soul out of affliction, and in Thy mercy shalt Thou utterly destroy mine enemies. And Thou shalt cut off all them that afflict my soul, for I am Thy servant.

Hearken unto me, O Lord, in Thy righteousness, and enter not into judgment with Thy servant. \twice\

Thy good Spirit shall lead me in the land of uprightness.

\glory\ \bothnow\

\alleluia\ \twice\

\alleluiaFinal

\note{Great Ektenia}

\choir\ Amen.

\deaconPriest\ In the \_\_ Tone: God is the Lord and hath appeared unto us. Blessed is he that cometh in the name of the Lord.

\stikh\ O give thanks unto the Lord, for He is good, for His mercy endureth forever.

\choir\ God is the Lord and hath appeared unto us. Blessed is he that cometh in the name of the Lord.

\stikh\ Surrounding me they compassed me, and by the name of the Lord I warded them off.

\choir\ God is the Lord and hath appeared unto us. Blessed is he that cometh in the name of the Lord.

\stikh\ I shall not die, but live, and I shall tell of the works of the Lord.

\choir\ God is the Lord and hath appeared unto us. Blessed is he that cometh in the name of the Lord.

\stikh\ The stone which the builders rejected, the same is become the head of the corner. This is the Lord’s doing, and it is marvelous in our eyes.

\choir\ God is the Lord and hath appeared unto us. Blessed is he that cometh in the name of the Lord.

\note{Then the troparia. On Sundays, the Resurrectional troparion is first, followed by troparia for the feast or saints, followed by the Resurrectional Theotokion in the same tone as the last troparion. A ``Glory'' is inserted before the second-to-last troparion, and ``Both now'' is inserted before the Theotokion.}

\centernote{Resurrectional Troparia}

\lbl{Tone 1} When the stone had been sealed by the Jews and the soldiers were guarding Thine immaculate Body, then didst Thou arise on the third, O Savior, granting life unto the world. Wherefore the powers of the heavens cried out to Thee, O Life-Giver: Glory to the Thy resurrection, O Christ; glory to Thy kingdom; glory to Thy dispensation, O only Friend of man.

\lbl{Tone 2} When Thou didst descend unto death, O Life immortal, then didst Thou slay Hades with the lightning of Thy divinity, and when Thou didst also raise the dead out of the nethermost depths all the power in the heavens cried out: O Life-Giver, Christ our God, glory be to Thee.

\lbl{Tone 3} Let the heavens rejoice, let earthly things be glad, for the Lord hath wrought might with His arm; He hath trampled upon death by death. The first-born of the dead hath He become. From the belly of Hades hath He delivered us, and hath granted great mercy to the world.

\lbl{Tone 4} Having learned the joyful proclamation of the resurrection from the angel, and having cast off the ancestral condemnation, the women disciples of the Lord spake to the Apostles exultantly: Death is despoiled and Christ God is risen, granting great mercy to the world.

\lbl{Tone 5} Let us, the faithful, praise and worship the Word, Who is co-beginningless with the Father and the Spirit, and Who was born of the Virgin for our salvation. For He was pleased to ascend the Cross in the flesh, and to endure death, and to raise the dead by His glorious resurrection.

\lbl{Tone 6} Angelic powers were above Thy tomb, and they that guarded Thee became as dead. And Mary stood by the grave seeking Thine immaculate body. Thou didst despoil Hades and wast not tried thereby. Thou didst meet the Virgin and didst grant us life. O Thou Who didst arise from the dead, Lord, glory be to Thee.

\lbl{Tone 7} Thou didst abolish death by the Cross. Thou didst open Paradise to the thief. Thou didst transform the myrrh-bearers lamentation and didst bid Thine Apostles to preach that Thou art risen, O Christ God, granting great mercy to the world.

\lbl{Tone 8} From on high didst Thou descend, O compassionate One. To burial of three days hast Thou submitted, that Thou mightest free us from our passions, O our Life and Resurrection, Lord, glory be to Thee.

\centernote{Resurrectional Theotokia}

\lbl{Tone 1} Whilst Gabriel was saying ``Rejoice'' to thee, O Virgin, at his voice the Master of all became incarnate in thee, the holy ark, as spake the righteous David. Thou wast seen to be more spacious than the heavens whilst holding thy Creator. Glory to Him that dwelt in thee; glory to Him that came forth from thee; glory to Him that hath delivered us through thy childbirth.

\lbl{Tone 2} All thy mysteries are beyond comprehension, all are exceedingly glorious, O Theotokos. Sealed with purity and preserved by virginity, thou wast shown to be a true mother, giving birth unto the true God. Do thou entreat Him that our souls be saved.

\lbl{Tone 3} We praise thee, the mediatress for the salvation of our race, O Virgin Theotokos. For in the flesh taken from thee, thy Son and our God hath deign to endure the passion through the Cross, and hath redeemed us from corruption, since He is the Friend of man.

\lbl{Tone 4} The mystery hidden from eternity and unknown to the angels is made manifest through, O Theotokos, to those on earth. God became incarnate by a union without conjunction and for our sake hath submitted willingly to the Cross, whereby He hath raised up the first-fashioned man, and hath saved our souls from death.

\lbl{Tone 5} Rejoice, O unentered gate of the Lord; rejoice, O battlement and shelter for them that flee to thee. Rejoice, O untroubled haven, thou who hast not know wedlock, who didst give birth in the flesh to thy Creator and God. Cease not to intercede for them who praise and worship Him to Whom thou gavest birth.

\lbl{Tone 6}

\lbl{Tone 7}

\lbl{Tone 8} O Thou who for our sakes was born of a virgin and didst suffer crucifixion, O Good One, and didst despoil death by death, and as God didst reveal the resurrection. Disdain them now which Thou has fashioned with Thy hand; show forth Thy love for mankind, O merciful One. Accept the Theotokos who gave Thee, who intercedeth for us, and do Thou, O our Savior, save a despairing people.

\centernote{Kathismata \& Sessional Hymns}

\note{Two kathismata from the Psalter are read, in between which are ektenia and sessional hymns. Consult the back of the Psalter to determine which kathismata should be read, depending on the time of year and the day of the week.}

\note{On Sundays, a method developed by St. John of Shanghai and San Francisco can be used instead:}

{\noindent
\begin{tabular}{l | c | c}
  Sunday Tone & First Psalm & Second Psalm \\
  Tone 1 & 9  & 17 (1-30) \\
  Tone 2 & 10 & 17 (31-46) \\
  Tone 3 & 11 & 18 \\
  Tone 4 & 12 & 19 \\
  Tone 5 & 13 & 20 \\
  Tone 6 & 14 & 21 \\
  Tone 7 & 15 & 22 \\
  Tone 8 & 16 & 23
\end{tabular}
}

\choir\ \lhm\ \thrice\

\glory\

\reader\ \bothnow\

\note{First appointed kathisma. On Sundays, this is Kathisma II.}

\note{Second appointed kathisma. On Sundays, this is Kathisma III.}

\end{document}
